Learning from Data Through Models

### 2023-09-04 Paper/draft.Rmd
editing

@@

@@

$$\frac{p[h^t|\tilde{\theta}}{p[h^t|\hat{\theta}]} > \alpha$$

@@

@@

@@

@@

>>>>

# Related Literature

@@

EXPLAIN WHY THEY ARE ALWAYS DISAPPOINTED AND UPDATE DOWNWARDS ON OMEGA UNTIL THEY FALL INTO THE TRAP.

@@

$$\frac{p[h^t|\tilde{\theta}]}{p[h^t|\hat{\theta}]} > \alpha$$

@@

EXPLAIN HOW THEY MIGHT GET OUT OF THE TRAP.

@@

ILLUSTRATE THE DIFFERENT PATHS WITH THE SIMULATION.

@@

The answers of part 2 allow me to classify subjects into overconfident, underconfident and correctly specified. If a subject 
guesses their score to be in a higher category than their true score, they are overconfident. On the contrary, if they guess their
score to be in a lower category than their true score, they are underconfident. Finally, if they guess their score to be in the same
category as their true score, they are correctly specified. This classification is done for each of the 6 topics separately.

@@

The randomness is controlled throughout the experiment and sessions by setting a seed drawn randomly before the
first session. By doing this I ensure that any two subjects who have the same type and face the same exogenous rate will observe
the same outcomes and thus, if they use the same updating procedure, they should be choosing the same gambles. This design feature allows me to identify differences in updating procedures across subjects.

### 2023-09-04 Paper/draft.tex
editing

\title{Learning from Data Through Models}
\author{Jimena Galindo}
\date{September 03, 2023}

@@

An agent is of type \(\theta \in \{\theta_L, \theta_M, \theta_H\}\) with
\(\theta_H > \theta_M > \theta_L\) and they face an unknown exogenous
state \(\omega \in \{\omega_L, \omega_M, \omega_H\}\) with
\(\omega_H>\omega_M>\omega_L\). Each of the values of \(\omega\) is
realized with equal probability. The agent knows the distribution of
\(\omega\) but not its value. They also hold some prior belief about
\(\theta\) which is potentially
misspecified.\footnote{The particular types of misspecification
that are allowed are discussed below.} The agent chooses a binary gamble
\(e in \{e_L, e_M, e_H\}\) and observes whether the gamble is a success
or a failure. ' If it is a success, the agent gets a payoff of \(1\) and
if it is a failure, the agent gets a payoff of \(0\). The probability of
success is increasing in both \(\theta\) and \(\omega\) and is fully
described in the following table:

  \multicolumn{1}{c}{} & \multicolumn{1}{c}{$\omega_H$} & \multicolumn{1}{c}{$\omega_M$} & \multicolumn{1}{c}{$\omega_L$}\\
  \cline{2-4}
  $e_H$ & 50 & 20 & 2 \\
  \cline{2-4}
  $e_M$ & 45 & 30 & 7 \\
  \cline{2-4}
  $e_L$ & 40 & 25 & 20 \\
  \cline{2-4}
  \multicolumn{1}{c}{} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{$\theta_L$} & \multicolumn{1}{c}{}\\
\end{tabular}
\hspace{.3cm} 
\begin{tabular}{ c|c|c|c|}

Conditional on a type, the agent's flow payoff is maximized by choosing
the gamble that matches the state. For example, if the value of
\(\omega\) is \(\omega_H\), the agent's flow payoff is maximized by
choosing \(e_H\) and if the state is \(\omega_L\) the flow payoff is
maximized by choosing gamble \(e_L\). The agent myopically chooses
gambles every period to maximize the flow payoff for \(T<\infty\)
periods, and the agent's payoff is the sum of the payoffs from each
period.

After observing the outcome of each gamble, the agent updates their
beliefs of both theta and omega in one of 4 different ways: As a
\emph{dogmatic} as in \citet{Heidhues2018} ; as a \emph{switcher} as in
\citet{Ba2023} ; as a \emph{self-attribution} updater as in
\citet{Coutts2020}; or as a fully Bayesian agent. Each of these updating
rules is described in detail below.

Notice that both \(\theta\) and \(\omega\) can be identified from the
outcomes if enough variation in the effort choices exists. This can be
seen by confirming that no two columns are the same. Therefore, the
agent can learn both their type and the state by observing enough
outcomes of the gambles for different effort choices.

@@

\[\frac{p[h^t|\tilde{\theta}]}{p[h^t|\hat{\theta}]} > \alpha\]

@@

@@

>>>>

\title{Learning from Data Through Models}
\author{Jimena Galindo}
\date{September 04, 2023}

@@

@@

\[\frac{p[h^t|\tilde{\theta}}{p[h^t|\hat{\theta}]} > \alpha\]

@@

EXPLAIN HOW THEY MIGHT GET OUT OF THE TRAP.

\hypertarget{a-simple-example}{%
\section{A Simple Example}\label{a-simple-example}}

The agent can be of one of 3 types:
\(\theta \in \{\theta_L, \theta_M, \theta_H\}\) with
\(\theta_H > \theta_M > \theta_L\). They face an unknown exogenous
success rate \(\omega \in \{\omega_L, \omega_M, \omega_H\}\) with
\(\omega_H>\omega_M>\omega_L\). Each of the values of \(\omega\) is
realized with equal probability. The agent knows the distribution of
\(\omega\) but not its realized value. Let the true type be \(\theta^*\)
and the true state be \(\omega^*\).

The agent holds some prior belief about \(\theta\)
\footnote{which is potentially misspecified as in the dogmatic and switcher cases discussed}
and chooses a binary gamble \(e in \{e_L, e_M, e_H\}\). The agent
observes whether the gamble is a success or a failure and gets a payoff
of \(1\) and if it is a success. They get \(0\) otherwise.

The probability of success is increasing in both \(\theta\) and
\(\omega\) and is fully described in the following table:

  \multicolumn{1}{c}{} & \multicolumn{1}{c}{$\omega_H$} & \multicolumn{1}{c}{$\omega_M$} & \multicolumn{1}{c}{$\omega_L$}\\
  \cline{2-4}
  $e_H$ & 50 & 20 & 2 \\
  \cline{2-4}
  $e_M$ & 45 & 30 & 7 \\
  \cline{2-4}
  $e_L$ & 40 & 25 & 20 \\

  \cline{2-4}
  \multicolumn{1}{c}{} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{$\theta_L$} & \multicolumn{1}{c}{}\\
\end{tabular}
\hspace{.3cm} 
\begin{tabular}{ c|c|c|c|}

Conditional on a type, the agent's flow payoff is maximized by choosing
the gamble that matches the state. For example, if the value of
\(\omega\) is \(\omega_H\), the agent's flow payoff is maximized by
choosing \(e_H\) and if the state is \(\omega_L\) the flow payoff is
maximized by choosing gamble \(e_L\). The agent myopically chooses
gambles every period to maximize the flow payoff for \(T\) periods.

After observing the outcome of each gamble, the agent updates their
beliefs using some algorithm and moves on to the next period.

Notice that both \(\theta\) and \(\omega\) can be identified from the
outcomes if enough variation in the effort choices exists. This can be
seen by confirming there is no pair of \(\theta\) and \(\omega\) such
that the probability of success is the same for all effort choices.
Therefore, by changing the effort choice, the agent can learn both their
type and the state if they observe enough outcomes.

In this example, for an agent with a dogmatic belief about their type, a
self-defeating equilibrium is one in which the agent chooses an effort
level that, under the true \(\theta\), yields a frequency of success
that is consistent with the agent's misspecified belief. That is, the
observed rate of success under the true type and the effort level
chosen, \(P[\text{sucess}|\theta^*, e]\) is the same as the probability
of success under the agent's belief,
\(P[\text{sucess}|\hat{\theta}, e]\).

In the data-generating process described above, there are six such
equilibria. For example, if the agent is of type \(\theta_M\) but
mistakenly believes that he is of type \(\hat{\theta}=\theta_H\) and the
and \(\omega^* = \omega_M\), when the effort chosen is \(e_L\), the
agent will observe a success with 45\% chance. Because the agent
dogmatically believes that their type is high, they will erroneously
conclude that the rate is \(\omega_L\). Under this belief, the optimal
action is \(e_L\) which will continue to generate successes with \(45%
\) probability, further reinforcing the incorrect belief. If the agent
were to learn the true type and rate, they would instead choose \(e_M\),
yielding a success with \(65%
\) probability and strictly higher expected payoffs.

From \citet{Heidhues2018}, we know that if \(\theta^*\) and \(\omega^*\)
are such that there is a self-confirming equilibrium, a dogmatic agent
will always fall into the trap and continue to make sub-optimal choices.
On the other hand, a switcher will be able to learn the true type and
state as long as their prior is diffused enough and their switching
threshold is low enough as in \citet{Ba2023}.

By including self-confirming equilibria, the example captures the forces
from each of the updating mechanisms discussed in the previous section
and allows for direct comparison of all the theories. For realizations
of \((\theta, \omega)\) for which there are self-confirming equilibria,
the dogmatic agent will fall into the trap whereas the switcher will be
able to escape it. Similarly, an agent with self-attribution bias will
update their beliefs differently from an unbiased Bayesian, leading them
to choose different gambles. I exploit such cases in order to test which
model is a better fit for how subjects behave in a laboratory
experiment.

ILLUSTRATE THE DIFFERENT PATHS WITH THE SIMULATION.

@@

I recruited XXX undergraduate subjects from the CESS lab at NYU who
participated in an in-person experiment. Sessions lasted approximately
XXX hours and subjects earned an average payment of XXX. The experiment
was programmed using oTree \citep{chen2016otree}.

The experiment consisted of 2 treatments: the ego-relevant condition and
the stereotype condition. Subjects participated in only one of them.
Treatments were randomly assigned at the session level. The tasks were
identical across treatments, but in the ego condition the probability of
success was based on the subject's own performance in a quiz, while in
the stereotype condition, the probability od success was based on the
performance of a randomly selected subject from another session.

The experiment had 3 parts. In Part 1 subjects had 2 minutes to answer
as many multiple-choice questions as they could from a 20 question quiz
in a particular topic. They did this for 6 different topics, one after
the other. The topics were: Math, Verbal Reasoning, Pop-culture and Art,
Science and Technology, US Geography, and Sports and Video Games. The
maximum number of questions they could answer correctly was 20. However,
they did not know how many questions were available and they were given
no feedback.

After taking all 6 quizzes, they proceeded to part 2 where they were
asked to guess their score on each of them. In the stereotype treatment
they were additionally asked to guess the score of another participant.
All they knew about the other participant was their gender identity and
whether they were US nationals or not. For each guess they had three
score options: 5 or fewer, between 6 and 15, 16 or more. Each of these
correspond to \(\theta_L\), \(\theta_M\), and \(\theta_H\) respectively.
They were also asked to say how confident they felt about their choices
with 4 options: it was a random guess, there is another equally likely
score, I am pretty sure, I am completely sure. These 4 answers are
mapped to priors that place probabilities .33, .50, .75, and 1 to the
chosen type. The remaining probability is split equally among the other
two types. Questions in part 2 were not incentivized, but subjects were
told that providing an accurate answer would increase their chances of
earning more money in the last part of the experiment.

The answers of part 2 allow me to classify subjects into overconfident,
underconfident and correctly specified. If a subject guesses their score
to be in a higher category than their true score, they are
overconfident. On the contrary, if they guess their score to be in a
lower category than their true score, they are underconfident. Finally,
if they guess their score to be in the same category as their true
score, they are correctly specified. This classification is done for
each of the 6 topics separately.

In part 3 subjects completed a belief updating task for each of the
quizzes. Before starting the task they were reminded of their guess for
the score. In the ego treatment they were reminded of their guess about
themselves and in the stereotype treatment they were reminded of their
guess about the other participant. In the stereotype treatment, they
were also reminded of the characteristics of the other participant.

For one topic at a time and in random order, they were presented with
the three gambles from the example and were asked to choose one of them.
The probability of success was determined by their own score in the
ego-relevant condition and by the score of the other participant in the
stereotype condition. Subjects had access to the three probability
tables in the printout on the instructions at all times and the meaning
of each cell was explained in detail in the instructions. However, on
the screen they had to choose which of the 3 tables they wanted to see
before entering their choice in it. This was done as an alternative to a
belief elicitation in each round. I take their choice of table to be a
noisy measure of their beliefs about the underlying type.

Once they have entered their choice, they observe a sample of 10
outcomes from the gamble they chose. After observing the outcomes, they
returned to the choice screen and entered a new choice. In the choice
screen subjects had access to the entire history of gambles and outcomes
for that task. Once they entered 11 gambles (and observed 110 outcomes),
they moved on to the next topic and repeated the same procedure. They
all did so for all 6 topics.

At the end of the experiment, one of the 6 topics was randomly selected
to determine the payment. They earned \(\$.20\) for each correct answer
in the quiz and for each success in part 3.

The randomness is controlled throughout the experiment and sessions by
setting a seed drawn randomly before the first session. By doing this I
ensure that any two subjects who have the same type and face the same
exogenous rate will observe the same outcomes and thus, if they use the
same updating procedure, they should be choosing the same gambles. This
design feature allows me to identify differences in updating procedures
across subjects.
